Sea-ice and macrozooplankton distribution as determinants of top predator community structure in Antarctic winter

### 2024-06-07 analysis/paper/paper.qmd
Table 2

```{r}
#| label: load
#| echo: false

@@

We surveyed the South Shetland Island and Elephant Island regions during the austral winters (August to September) of 2012–2016 as part of the US Antarctic Marine Living Resources (AMLR) program aboard the RV Ice Breaker (RVIB) ‘Nathaniel B. Palmer’ [@fig-overviewmap]. The main objectives of the survey were to map hydrographic conditions, study the distribution and abundance of fish, krill, and other zooplankton through acoustic methods and net hauls, and describe the spatial distribution of seabirds and marine mammals. We followed methods consistent with the long-term austral summer surveys conducted by the US AMLR program [@reissOverwinterHabitatSelection2017]. We sampled between 33 and 108 stations per year, dependent on ice conditions. Stations were spaced approximately 25–30 km apart, and the full survey area covered an area of approximately 120000 km^2^.

@@

During the 5 survey years, we sighted a total of 22 species of seabirds and marine mammals (Table 2). The most abundant species were snow petrels, Adélie penguins, and Antarctic fur seals. The most widespread species observed were snow petrels, Antarctic petrels, and southern giant petrels. The 6 most abundant species (total individuals observed in Table 2) averaged across years were widely distributed across the study region, although densities varied by species [@fig-top6map]. Distribution maps of all seabird (Fig. S1) and marine mammal species (Fig. S2) are provided in the Supplement. We observed predators in proximity (15 km, 3 d) to 245 sampling stations, ranging from 19 stations in 2012 to 68 stations in 2014. Each station was associated with 30.2 ± 10.3 km (mean ± SD) of visual survey effort (Table S1).

@@

We retained 3 predator clusters from the hierarchical analysis, which we refer to as the ‘pack ice’, ‘marginal ice’, and ‘open water’ communities based on their indicator species (Table 3) and relationships to sea-ice coverage [@fig-seaice]. See Fig. S3 for the gap statistic curve, which we used to identify the optimal number of clusters.

@@

The open water community was observed most frequently (56.3% of sites), followed by the marginal ice (22.0%) and pack ice (21.6%) communities (Table 4), with variability between years related to the spatial distribution of sea ice [@fig-predclustannual]. Geographically, in general, the open water community occupied the northern offshore regions and the gap between Elephant Island and the rest of the Shetland Islands [@fig-predclustkde]. We found predators characteristic of the marginal ice community in the western Bransfield Strait and north of Elephant Island. The pack ice community primarily occupied the eastern Bransfield Strait.

@@

We retained 3 NMDS axes for investigating the relationships between the environment and community structure of predators (stress = 0.135). Two variables, sampling year and ice coverage, were significantly correlated with the ordination scores (p = 0.002 and 0.013, respectively), although the effect size was small (r^2^ = 0.061 and 0.054, respectively; Table 3; see Table S3 for the coefficient, r^2^, and p-values of the environmental fit regression). The pack ice cluster occupied a wide range of values on NMDS axis 1, overlapping with the other 2 clusters [@fig-nmds]. The marginal ice and open water clusters were well separated by NMDS axis 2.

@@

We analyzed the composition and distribution of predator communities around the northern Antarctic Peninsula in winter to investigate associations among marine predators, their prey, and the physical environment. We identified 3 communities of predators, and hypothesized that the spatial organization of these communities is related to the distribution of macrozooplankton communities and a gradient in sea-ice coverage. The pagophilic community (e.g. Adélie penguins, crabeater seals) consisted of species with diets dominated by Antarctic krill (Table 2) that rely on the pack ice throughout the annual cycle [@ainleyBirdsAntarcticSea2017; @besterAntarcticMarineMammals2017]. Predators in the marginal ice zone (e.g. Antarctic fur seals, snow petrels) and in open water (e.g. southern fulmars, other petrels) communities also consume Antarctic krill, but their diets are more diverse, with a greater reliance on fish, especially myctophids such as *Electrona antarctica* (Table 2) [@ridoux1989diets; @ainley1992does; @lorentsen1998diet]. Despite the diet overlap between these 2 communities, they tend to segregate spatially according to ice coverage due to differing foraging strategies. Marginal ice zone predators use ambush and pursuit plunging behavior to exploit prey among ice floes, which open water predators have not been observed to do [@ainleySpeciesHabitatRelationshipsAntarctic1993].

@@

Our data represent a baseline of biophysical conditions during a period of relatively high sea-ice coverage compared to recent declines [@eayrsRapidDeclineAntarctic2021]. Our 5 yr winter synthesis of sea ice, hydrography, macrozooplankton, and top predators is unlikely to be replicated in the near future, due to increased costs of shipboard research, reduced availability of research vessels, and prioritized sampling during the Antarctic summer. The spatial averages of community composition (i.e. predator clusters; @fig-predclustkde) may be considered essential foraging habitats for overwintering predators. The open water assemblage occupies the most temporally variable ocean habitat that we sampled, which changes based on currents and latitudinal position of the marginal ice zone. The marginal ice habitat was considerably smaller, with 2 important concentrations: southwest part of the Bransfield Strait, and open water to the north of Elephant Island. Antarctic krill are consistently concentrated in the southern region over multiple years [@reissOverwinterHabitatSelection2017], and stable sea-ice concentrations are more likely to occur there due to coastal ocean processes (coastal currents, formation of eddies) and shifts in ice from higher latitudes [@moffatShelfOceanExchange2018]. We found consistently high concentrations of crabeater seals and fur seals within both marginal ice zone habitats [@fig-top6map; @fig-predclustannual], and the presence of sea ice likely contributes by supporting haul-out habitat for seal thermoregulation and rest between foraging bouts. There were more concentrations of fur seals compared to crabeater seals within the marginal ice zone north of Elephant Island. We did not always encounter high concentrations of krill in this region, which might explain the reduction in crabeater seals, or that fur seals were targeting other forage taxa, perhaps myctophids [@daneriDietAntarcticFur2005; @politoInvestigatingUseStable2010; @descalzoRelevanceFishSummer2023], that our net hauls were inadequate at catching.

@@

>>>>

```{r}
#| label: load

@@

We surveyed the South Shetland Island and Elephant Island regions during the austral winters (August to September) of 2012–2016 as part of the US Antarctic Marine Living Resources (AMLR) program aboard the RV Ice Breaker (RVIB) ‘Nathaniel B. Palmer’ ([@fig-overviewmap]). The main objectives of the survey were to map hydrographic conditions, study the distribution and abundance of fish, krill, and other zooplankton through acoustic methods and net hauls, and describe the spatial distribution of seabirds and marine mammals. We followed methods consistent with the long-term austral summer surveys conducted by the US AMLR program [@reissOverwinterHabitatSelection2017]. We sampled between 33 and 108 stations per year, dependent on ice conditions. Stations were spaced approximately 25–30 km apart, and the full survey area covered an area of approximately 120000 km^2^.

@@

During the 5 survey years, we sighted a total of 22 species of seabirds and marine mammals ([@tbl-predators]). The most abundant species were snow petrels, Adélie penguins, and Antarctic fur seals. The most widespread species observed were snow petrels, Antarctic petrels, and southern giant petrels. The 6 most abundant species (total individuals observed in @tbl-predators) averaged across years were widely distributed across the study region, although densities varied by species ([@fig-top6map]). Distribution maps of all seabird (Fig. S1) and marine mammal species (Fig. S2) are provided in the Supplement. We observed predators in proximity (15 km, 3 d) to 245 sampling stations, ranging from 19 stations in 2012 to 68 stations in 2014. Each station was associated with 30.2 ± 10.3 km (mean ± SD) of visual survey effort (Table S1).

@@

We retained 3 predator clusters from the hierarchical analysis, which we refer to as the ‘pack ice’, ‘marginal ice’, and ‘open water’ communities based on their indicator species (Table 3) and relationships to sea-ice coverage ([@fig-seaice]). See Fig. S3 for the gap statistic curve, which we used to identify the optimal number of clusters.

@@

The open water community was observed most frequently (56.3% of sites), followed by the marginal ice (22.0%) and pack ice (21.6%) communities (Table 4), with variability between years related to the spatial distribution of sea ice ([@fig-predclustannual]). Geographically, in general, the open water community occupied the northern offshore regions and the gap between Elephant Island and the rest of the Shetland Islands ([@fig-predclustkde]). We found predators characteristic of the marginal ice community in the western Bransfield Strait and north of Elephant Island. The pack ice community primarily occupied the eastern Bransfield Strait.

@@

We retained 3 NMDS axes for investigating the relationships between the environment and community structure of predators (stress = 0.135). Two variables, sampling year and ice coverage, were significantly correlated with the ordination scores (p = 0.002 and 0.013, respectively), although the effect size was small (r^2^ = 0.061 and 0.054, respectively; Table 3; see Table S3 for the coefficient, r^2^, and p-values of the environmental fit regression). The pack ice cluster occupied a wide range of values on NMDS axis 1, overlapping with the other 2 clusters ([@fig-nmds]). The marginal ice and open water clusters were well separated by NMDS axis 2.

@@

We analyzed the composition and distribution of predator communities around the northern Antarctic Peninsula in winter to investigate associations among marine predators, their prey, and the physical environment. We identified 3 communities of predators, and hypothesized that the spatial organization of these communities is related to the distribution of macrozooplankton communities and a gradient in sea-ice coverage. The pagophilic community (e.g. Adélie penguins, crabeater seals) consisted of species with diets dominated by Antarctic krill (\[\@tbl-predators\]) that rely on the pack ice throughout the annual cycle [@ainleyBirdsAntarcticSea2017; @besterAntarcticMarineMammals2017]. Predators in the marginal ice zone (e.g. Antarctic fur seals, snow petrels) and in open water (e.g. southern fulmars, other petrels) communities also consume Antarctic krill, but their diets are more diverse, with a greater reliance on fish, especially myctophids such as *Electrona antarctica* (\[\@tbl-predators\]) [@ridoux1989diets; @ainley1992does; @lorentsen1998diet]. Despite the diet overlap between these 2 communities, they tend to segregate spatially according to ice coverage due to differing foraging strategies. Marginal ice zone predators use ambush and pursuit plunging behavior to exploit prey among ice floes, which open water predators have not been observed to do [@ainleySpeciesHabitatRelationshipsAntarctic1993].

@@

Our data represent a baseline of biophysical conditions during a period of relatively high sea-ice coverage compared to recent declines [@eayrsRapidDeclineAntarctic2021]. Our 5 yr winter synthesis of sea ice, hydrography, macrozooplankton, and top predators is unlikely to be replicated in the near future, due to increased costs of shipboard research, reduced availability of research vessels, and prioritized sampling during the Antarctic summer. The spatial averages of community composition (i.e. predator clusters; @fig-predclustkde) may be considered essential foraging habitats for overwintering predators. The open water assemblage occupies the most temporally variable ocean habitat that we sampled, which changes based on currents and latitudinal position of the marginal ice zone. The marginal ice habitat was considerably smaller, with 2 important concentrations: southwest part of the Bransfield Strait, and open water to the north of Elephant Island. Antarctic krill are consistently concentrated in the southern region over multiple years [@reissOverwinterHabitatSelection2017], and stable sea-ice concentrations are more likely to occur there due to coastal ocean processes (coastal currents, formation of eddies) and shifts in ice from higher latitudes [@moffatShelfOceanExchange2018]. We found consistently high concentrations of crabeater seals and fur seals within both marginal ice zone habitats ([@fig-top6map; @fig-predclustannual]), and the presence of sea ice likely contributes by supporting haul-out habitat for seal thermoregulation and rest between foraging bouts. There were more concentrations of fur seals compared to crabeater seals within the marginal ice zone north of Elephant Island. We did not always encounter high concentrations of krill in this region, which might explain the reduction in crabeater seals, or that fur seals were targeting other forage taxa, perhaps myctophids [@daneriDietAntarcticFur2005; @politoInvestigatingUseStable2010; @descalzoRelevanceFishSummer2023], that our net hauls were inadequate at catching.

@@

```{r}
#| label: tbl-predators
#| tbl-cap: "Predator species observed during surveys. Ind: total individuals observed, Sites: number of sites where species was present, Freq: percentage of sites where species was present. Typical prey and foraging behavior are presented for context."
#| tbl-colwidths: [18, 7, 7, 11, 17, 16, 24]

format_percent <- scales::label_percent(0.1)

predator_context <- tribble(
  ~species, ~`Typical prey`, ~`Foraging behavior`,     ~Reference,
  "SNPT",   "Myctophids",    "Surface feeding (<5 m)", "@spearMorphologicalDifferencesRelative1998; @delordSpeciesSpecificForaging2016",
  "ANPT",   "Myctophids, krill, squid",    "Surface feeding (<5 m)", "@spearMorphologicalDifferencesRelative1998; @delordAntarcticPetrelsIce2020",
  "SGPT",   "Juvenile pinnipeds and seabirds, carrion",    "Scavenging, surface feeding (<5 m)", "@hunterFoodFeedingEcology1983",
  "FUSE",   "Krill, myctophids",    "Diving (~15-50 m)", "@stanilandForagingBehaviourTwo2011",
  "KEGU",   "Limpets, krill",    "Surface feeding (<5 m)", "@silvaDoesAccessHighquality2001",
  "ANFU",   "Myctophids, krill, squid",    "Surface feeding (<5 m)", "@ridoux1989diets; @spearMorphologicalDifferencesRelative1998",
  "CAPT",   "Krill, myctophids",    "Surface feeding (<5 m)", "@ridoux1989diets; @spearMorphologicalDifferencesRelative1998",
  "ADPN",   "Krill",    "Diving (~15-50 m)", "@ciminoClimatedrivenSympatryMay2016; @juaresDietAdeliePenguins2018",
  "CRSE",   "Krill",    "Diving (~50-100 m)", "@burnsWinterHabitatUse2004; @huckstadtDietSpecialistChanging2012",
  "BLPT",   "Krill, myctophids, amphipods",    "Surface feeding (<5 m)", "@princeFoodFeedingEcology1980",
  "LESE",   "Seals, penguins, fish, rkill",    "Diving (~50-100 m)", "@walkerSeasonalOccurrenceDiet1998; @casauxDietLeopardSeal2009; @krauseNovelForagingStrategies2015",
  "ANTE",   "Krill, myctophids",    "Surface feeding (<5 m)", "@croxallFoodFeedingEcology1980; @ainley1992does",
  "PFSB",   "Carrion",    "Scavenging", "@faveroForagingEcologyPaleFaced1996",
  "MIWH",   "Krill",    "Diving (~5-100 m)", "@friedlaenderFeedingRatesUnderice2014",
  "WESE",   "Fish, squid",    "Diving (~100-350 m)", "@testaOverwinterMovementsDiving1994; @lakeRegionalTemporalFinescale2003",
  "GEPN",   "Krill, fish",    "Diving (~25-100 m)", "@croxallFoodFeedingEcology1980; @ciminoClimatedrivenSympatryMay2016",
  "ELSE",   "Myctophids, squid",    "Diving (~200-800 m)", "@guinetSouthernElephantSeal2014; @daneriFeedingHabitsSouthern2015",
  "KIWH",   "Whales, seals, toothfish, penguins",    "Complex and diverse", "@pitmanThreeFormsKiller2003",
  "EMPN",   "Krill, silverfish",    "Diving (~50-300 m)", "@kirkwoodForagingEcologyFemale1997; @rodaryBenthicDivingMale2000",
  "SBWH",   "Squid",    "Diving (>500 m)", "@macleodReviewDataDiets2003",
  "ANSH",   "Demersal fish",    "Diving (<50 m)", "@casauxShagsAntarcticaTheir2006",
  "ROSE",   "Myctophids, squid",    "Diving (~100-300 m)", "@blixRossSealOmmatophoca2007; @southwellReviewDataAbundance2012"
)

n_sites <- nrow(read_obj("zoop_sf"))

read_obj("predators_stations") %>% 
  mutate(common = code_to_common(species),
         scientific = code_to_scientific(species),
         Species = str_glue("{common} _{scientific}_")) %>% 
  group_by(Species, species) %>% 
  summarize(Ind = sum(count),
            Sites = n(),
            `Freq (%)` = format_percent(Sites / n_sites[1]),
            .groups = "drop") %>% 
  left_join(predator_context, by = "species") %>% 
  select(-species) %>% 
  arrange(desc(Sites)) %>% 
  knitr::kable()
